The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-07-05 ms/main.tex
intermediate backup because the computer is heating up

@@

The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is
%
\begin{equation}\label{eq:dispersal}
\begin{aligned}
m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
\end{aligned}
\end{equation}
where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches. The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}

@@

Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents. We also consider that mutants are rare compared to wild-type individuals in old-habitat patches. Denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of successful offspring of mutant individuals in this old-habitat patch (i.e., of offspring that survive density regulation and become adults in the next generation) is 
%
\begin{equation}\label{eq:sold}
K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
%
\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
%
= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
\end{equation}
%
The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches. Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$. For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$.

\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal. Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough. In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation. Competition is fiercer as carrying capacity is nearer. }\footnote{\florence{I think this is more Results than Model}}. \linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI); formulas for $a_{\text{old}}$ are given in eq.~\ref{Seq:s_old}.

@@

\linelabel{R1-24}The survival probability of this multi-type branching process, $\varphi_{k}$ with $k$ indicating the initial habitat type of the mutant, is then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}
\begin{equation}\label{eq:ext_prob}
	\begin{aligned}
		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{ offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
		&=\exp\left[-\left(1-m^{\text{old}\to \text{new}}_m\right)(1+a_{\text{old}})\varphi_{\text{old}} - m^{\text{old}\to \text{new}}_m (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, ,   \\
		1-\varphi_{\text{new}} &= \exp\left[-m^{\text{new}\to \text{old}}_m(1+a_{\text{old}})\varphi_{\text{old}} - \left(1-m^{\text{new}\to \text{old}}_m\right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
	\end{aligned}
\end{equation}
Intuitively the left hand side, the extinction probability, equals the sum over all possible scenarios of the trajectories towards extinction, e.g. the initial individual having $j$ offspring in a patch of type $k$, old or new, and all of these $j$ offspring becoming extinct ($(1-\varphi_{\text{k}})^j$). Unfortunately, a general analytical solution to these equations is not accessible, but they can be solved numerically. For weak selection and (potentially) weak dispersal -- i.e. $a_{\text{old}},a_{\text{new}},m\ll 1$ needs to hold for at least two of the three parameters -- an approximation is available: see for example~\citet[Theorem~5.6]{haccou_book} for the general theory and \citet{tomasini_2018} for an application in a similar setting. The detailed derivation is presented in the SI, Section~S2. We find
\begin{equation}\label{eq:estab_approx}
	\begin{aligned}
		\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}}&+ & \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
		& & &  \; \; \; + \; \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
		\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} &+ & \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
	& & &  \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, ,}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
	\end{aligned}
\end{equation}
where $C$ is a scaling constant that depends on $m,f_{\text{old}},\widehat{\pi}_m,a_{\text{old}}$ and $a_{\text{new}}$ through
\begin{equation}\label{eq:normalization}
	C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, . 
\end{equation}

@@

\section{Deriving the model dynamics \label{sec:app:modeldyn}}
\linelabel{test}In this section we provide the mathematical \chg{details} of the model that is verbally described in the main text. We start by deriving the population dynamics of the wild type \chg{when alone} in both patch types. This will allow us to compute the local growth rate of the mutant \chg{when rare} in old-habitat patches, $a_{\text{old}}$.

Before we go into the details of the computation, we recall the form of the dispersal rates. For a wild-type individual to disperse to a new-habitat patch, this probability is given by

\begin{equation}\label{Seq:dispersal_rates_old}
m_w^{\text{new}} = m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, ,
\end{equation}
%
where $m$ denotes the emigration probability, $f_{\text{old}}$ the frequency of old-habitat patches and $\widehat{\pi}_w$ the transformed wild-type bias towards old-habitat patches.
Analogously, the probability for a wild-type individual \chg{to emigrate and} to move to an old-habitat patch reads
%
\begin{equation}\label{Seq:dispersal_rate_new}
m_w^{\text{old}} = m\, \frac{\widehat{\pi}_w f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, .
\end{equation}

@@

\subsection*{Stationary wild-type population sizes}
We derive the (deterministic) stationary population sizes of the wild-type in both patch types. These values are denoted by $\widehat{N}_w^{k}$, where $k$ corresponds to the patch type (old or new). In old-patch habitats we assume that the population is always at carrying capacity. Therefore, we have $\widehat{N}_w^{\text{old}} = K_{\text{old}}$.

In new-habitat patches, the stationary value is given by the solution of the following equation:

\begin{equation}\label{Seq:wt_deme2}
\begin{aligned}
\widehat{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)(1-r)\widehat{N}_w^{\text{new}} \\
&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} (1-r)  \widehat{N}_w^{\text{old}}\\
&= \left(1-m+m_w^{\text{new}} \right)(1-r)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \left(\frac{f_{\text{old}}}{1-f_{\text{old}}}\right) (1-r)  K_{\text{old}},
\end{aligned}
\end{equation}
%
which simplifies to
%
\begin{equation}
\widehat{N}_w^{\text{new}} = \frac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\ .
\end{equation}
%
Since we assume density regulation, this value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. Hence, we find
\begin{equation}\label{Seq:Nhatnew}
\widehat{N}_w^{\text{new}} =\left\{ \begin{array}{ll} 
K_{\text{new}}, & \text{if }\dfrac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\geq K_{\text{new}}; \\
%
\dfrac{m (1-r) f_{\text{old}} K_{\text{old}}
}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}, & \text{otherwise}.
\end{array}\right.
\end{equation}

\subsection*{Wild-type population sizes after dispersal}
In order to explicitly compute the growth rate of the mutant in old-habitat patches, we need an analytical expression for the number of wild-type individuals after the dispersal step. Later on, in the approximation of the probability of adaptation, we also use the number of wild-type individuals in new-habitat patches before reproduction. We denote these values by $\widetilde{N}_w^k$, where $k$ corresponds to the habitat type (old or new). These are given as the solutions\florence{solution sounds like there is something to solve, which does not seem to be the case. Am I missing something?} to the following equations

\begin{equation}
\begin{aligned}
\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \frac{\widehat{\pi}_w f_{\text{old}}}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})}\,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \\
\widetilde{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{1-f_{\text{old}}+\pi f_{\text{old}}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .
\end{aligned}
\end{equation}
Using $\widehat{N}_w^{\text{old}}=K_{\text{old}}$ and in the case of $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ this reduces to
\begin{equation}\label{Seq:wt_eq1}
\begin{aligned}
\widetilde{N}_w^{\text{old}} &= \frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \\ 
\widetilde{N}_w^{\text{new}}&= \frac{(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w (1-m))K_{\text{new}} + m f_{\text{old}} K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\ .
\end{aligned}
\end{equation}
%
If on the other hand $\widehat{N}_w^{\text{new}}<K_{\text{new}}$ holds, using eq.~\eqref{Seq:Nhatnew}\footnote{\florence{please check the eqref I wrote}}, we obtain\florence{separate into 2 eqs for more precise reference}
\begin{equation}\label{Seq:wt_eq2}
\begin{aligned}
\widetilde{N}_w^{\text{old}} &= \left(1-m+m_w^{\text{old}} \right)K_{\text{old}}\\
&\qquad \qquad + \left(m_w^{\text{old}} \frac{m (1-r) f_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\left(\frac{1-f_{\text{old}}}{f_{\text{old}}}\right)\right) K_{\text{old}} \\
%&= \left(1-m_1+\frac{m_1 m_2 (1-r)}{(1-(1-m_2)(1-r))}\right) K \\ 
&= \frac{m \widehat{\pi}_w f_{\text{old}} + r (1-m) (1-f_{\text{old}} (1-\widehat{\pi}_w))}{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} K_{\text{old}}\ \\
&= \left(1 - \frac{r m (1-f_{\text{old}}) }{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} \right) K_{\text{old}}\, ,\\
\widetilde{N}_w^{\text{new}} &= \left(1-m+m_w^{\text{new}}\right) \frac{m (1-r) f_{\text{old}} }{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\, K_{\text{old}}\\
&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}}}{(1-f_{\text{old}})} K_{\text{old}}\\
%&= \left(\frac{(1-m_2) m_1 (1-r) f }{(1-f)(1-\left(1-m_2 \right)(1-r))} + \frac{m_1 f}{(1-f)}\right) K\\
&= \frac{m f_{\text{old}}}{m \widehat{\pi}_w f_{\text{old}} + r(1- f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}  (1-m)) }\, K_{\text{old}}\, .
\end{aligned}
\end{equation}

\subsection*{Wild-type population sizes during the environmental change}
\chg{Lastly, compute the (deterministic) wild-type population size over time during the environmental change. This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.}

>>>>

%%% Fix for lineno and align
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
	\renewenvironment{#1}%
	{\linenomath\csname old#1\endcsname}%
	{\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
	\patchAmsMathEnvironmentForLineno{#1}%
	\patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
	\patchBothAmsMathEnvironmentsForLineno{equation}%
	\patchBothAmsMathEnvironmentsForLineno{align}%
	\patchBothAmsMathEnvironmentsForLineno{flalign}%
	\patchBothAmsMathEnvironmentsForLineno{alignat}%
	\patchBothAmsMathEnvironmentsForLineno{gather}%
	\patchBothAmsMathEnvironmentsForLineno{multline}%
}
%%%

@@

The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is\florence{TODO change into simple bias (remove m)}
%
\begin{equation}\label{eq:dispersal}
\begin{aligned}
m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
\end{aligned}
\end{equation}
\florence{there seems to be a problem with the second equation, should be $m -$ no? }
where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches. The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}

@@

A key assumption to our mathematical analysis is that the mutant individuals are rare enough that their dynamics do not affect the wild-type population, at least during the establishment phase of the mutant lineage. The wild-type population sets a demographic context that affects mutants dynamics. The mathematical analysis therefore focuses on the population dynamics of the mutant population, considering the wild-type population as constant over time.\footnote{\florence{please check this is correct}}

We assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents. Denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of successful offspring of mutant individuals in this old-habitat patch (i.e., of offspring that survive density regulation and become adults in the next generation) is 
%
\begin{equation}\label{eq:sold}
K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
%
\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
%
\overset{\mathrm{def}}{=} \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
\end{equation}
%
The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches. Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$. It depends on $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal, which is calculated in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI) (Eq.~\eqref{Seq:Ntildeold}).

\st{For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$. }\florence{com eback to this}

%\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal. Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough. In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation. Competition is fiercer as carrying capacity is nearer. }\footnote{\florence{I think this is more Results than Model}}. \linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI).

@@

\linelabel{R1-24}We denote by $\varphi_{\text{old}}$ (resp. $\varphi_{\text{new}}$) the probability of establishment of this two-type branching process when the mutant is initially located in a old- (resp. new-) habitat patch. This probability is computed by considering all possible ways of going extinct: the initial individual having $j$ successful offspring in a patch of type $k$, but all lineages descending from these $j$ offspring eventually go extinct; then summing over $k$ and $j$. The establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ are then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}:
%
\begin{equation}\label{eq:ext_prob}
	\begin{aligned}
		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ successful offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{successful offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
		&=\exp\left[-\left(1-m^{\text{old}\to \text{new}}_m\right)(1+a_{\text{old}})\varphi_{\text{old}} - m^{\text{old}\to \text{new}}_m (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, ,   \\
		1-\varphi_{\text{new}} &= \exp\left[-m^{\text{new}\to \text{old}}_m(1+a_{\text{old}})\varphi_{\text{old}} - \left(1-m^{\text{new}\to \text{old}}_m\right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
	\end{aligned}
\end{equation}
This system of equations can be solved numerically. An analytical approximate solution is available in the case of weak selection and (potentially) weak dispersal (i.e. $a_{\text{old}},a_{\text{new}},m\ll 1$ needs to hold for at least two of the three parameters); see for example~\citet[Theorem~5.6]{haccou_book} for the general theory and \citet{tomasini_2018} for an application in a similar setting. The detailed derivation is presented in the SI, Section~\ref{sec:approxestabproba}. We find
\begin{equation}\label{eq:estab_approx}
	\begin{aligned}
		\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}}&+ & \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
		& & &  \; \; \; + \; \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
		\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} &+ & \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
	& & &  \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, ,}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
	\end{aligned}
\end{equation}
where $C$ is a scaling constant that depends on $m,f_{\text{old}},\widehat{\pi}_m,a_{\text{old}}$ and $a_{\text{new}}$ through
\begin{equation}\label{eq:normalization}
	C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, . 
\end{equation}

@@

\section{Deriving the model dynamics \label{sec:app:modeldyn}}
\linelabel{test}In this section we provide the mathematical details of the model that is verbally described in the main text. We start by deriving the population dynamics when only the wild type is present. This will allow us to compute the local growth rate of a rare mutant.

Before we go into the details of the computation, we recall the form of the dispersal rates. A dispersing wild-type individual immigrates to a new-habitat patch with probability

\begin{equation}\label{Seq:dispersal_rates_old}
\mu_w^{\text{new}} = \frac{1-f_{\text{old}}}{1-f_{\text{old}} + e^{\pi_w} f_{\text{old}}} = 1 - \mu_w^{\text{old}} \, ,
\end{equation}
%
where $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_w$ is the wild-type bias towards old-habitat patches. The complement, $\mu_w^{\text{old}}$, is the probability that the dispersing wild-type individual instead immigrates into an old-habitat patch.

@@

\subsection*{Stationary wild-type population sizes}
We denote by $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ the deterministic stationary population sizes of the wild type in old- and new-habitat patches. We assume that the population is always at carrying capacity in old-habitat patches, so that $\widehat{N}_w^{\text{old}} = K_{\text{old}}$. We now compute $\widehat{N}_w^{\text{new}}$ recursively: it is given by the solution of the following equation: 
%
\begin{equation}\label{Seq:wt_deme2}
\widehat{N}_w^{\text{new}} = \left(1-m + m \mu_w^{\text{new}}  \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)  \omega_w^{\text{new}}  \widehat{N}_w^{\text{new}}   + m \mu_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \omega_w^{\text{new}}  \widehat{N}_w^{\text{old}}, \nonumber 
\end{equation}
where the first term on the right-hand side corresponds to individuals born in a new-habitat patch and staying in it or migrating and landing in a new-habitat patch, and the second term corresponds to individuals born in an old-habitat patch and migrating to a new-habitat patch. Simplifying, using eq.~\eqref{Seq:dispersal_rates_old} to replace $\mu_w^{\text{new}} $ and replacing $\widehat{N}_w^{\text{old}}$ by $K_{\text{old}}$ , we obtain
%
\begin{equation}
\widehat{N}_w^{\text{new}} = \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
1-f_{\text{old}} + e^{\pi_w} f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + e^{\pi_w} f_{\text{old}} (1-m))
}
\ .
\end{equation}
%
This value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. So, 
\begin{equation}\label{Seq:Nhatnew}
\widehat{N}_w^{\text{new}} = \min\left(K_{\text{new}},  \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
	1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))
} \right).
\end{equation}

\subsection*{Wild-type population sizes after dispersal}

We denote by $\widetilde{N}_w^{\text{old}}$ and $\widetilde{N}_w^{\text{new}}$ the numbers of wild-type individuals \textit{after the dispersal step}. These quantities are needed to explicitly compute the growth rate of the mutant in old-habitat patches, and to approximate the probability of adaptation. They are given by
%
\begin{subequations}\label{Seq:Ntilde}
\begin{align}
\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \mu_w^{\text{old}} \,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m \mu_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \label{Seq:Ntildeold}\\
%
%
\widetilde{N}_w^{\text{new}} &= \left(1-m + m \mu_w^{\text{new}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m \mu_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .\label{Seq:Ntildenew}
\end{align}
\end{subequations}
%
We then replace $\widehat{N}_w^{\text{old}}$ by $K_{\text{old}}$ (since old-habitat patches are assumed to be at carrying capacity after density regulation), and $\widehat{N}_w^{\text{new}}$ by the formula given in eq.~\eqref{Seq:Nhatnew}.

\subsection*{Wild-type population sizes during the environmental change}
Lastly, compute the (deterministic) wild-type population size over time during the environmental change. This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.

At the moment a patch deteriorate, its population size is still given by the carrying capacity of the old habitat, $K_{\text{old}}$, but only each adult has now on average $\omega_{m}^{\text{new}}<1$ offspring. As the size of the local population decreases, there is not need for density regulation anymore. Neglecting dispersal for the moment, in generation $\tau$ after the degradation of a patch\footnote{\florence{is this just said/shown for the sake of pedagogy, or do you actually need this equation? If not, I suggest removing it.}} , we would have 
	\begin{equation}
	N_w^{\text{new}}(\tau) = K_{\text{old}}(\omega_w^{\text{new}})^{\tau}
	\end{equation}
	wild-type individuals in this patch. Including dispersal between the patches then results in the following number of wild-type individuals in a patch $i$ at time $\tau$ post-degradation, given that there are $k-1$ other new-habitat patches
	\begin{equation}
	N_w^{i,k}(\tau) = \omega_w^{\text{new}}\left((1-m)N_w^{i,k}(\tau-1) + \frac{m_w^{\text{new}}}{k} (M-k) K_{\text{old}} + \frac{m_w^{\text{new}}}{k} \sum_{l=1}^k N_w^{l,k}(\tau-1) \right)\ .
	\end{equation}
	The first term represents the remaining individuals after emigration, the second and third term are immigrants from old- and new-habitat patches (distributed equally among the $k$ new-habitat patches), respectively.

\subsection*{The local per capita growth rate $a_{\text{old}}$}
As stated in eq.~\eqref{eq:sold} in the main text, we define the per capita growth rate of rare mutant in the old habitat by
\begin{equation}
1 + a_{\text{old}} = K_{\text{old}}\, \dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}_w^{\text{old}}}\, ,    
\end{equation}
%
where we replace $\widetilde{N}_w^{\text{old}}$ by the formula given in \eqref{Seq:Ntildeold}. 
%Having computed the number of wild-type individuals after dispersal in an old-habitat patch, $\widetilde{N}_w^{\text{old}}$, we can write the growth rate as follows \florence{in which case would we have $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ ?}

%\begin{equation}\label{Seq:s_old}
%a_{\text{old}} = \left\{ \begin{array}{ll}
%\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
%\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}. 
%\end{array}
%\right.
%\end{equation}

\subsection*{The local per capita growth rate $a_{\text{new}}$}
In new-habitat patches, we assume that the carrying capacity $K_{\text{new}}$ is not reached during the establishment phase of the mutant. In other words, we assume that there is no density regulation in new-habitat patches. As a result, the local per capita growth rate of the mutant only depends on its fecundity: 
%
\begin{equation}
1 + a_{\text{new}} = \omega_m^{\text{new}}.
\end{equation}

%\chg{In new-habitat patches we assume that during the relevant phase of establishment of a rare mutant the carrying capacity $K_{\text{new}}$ is not reached\footnote{\florence{This seems at odds with the commented out sentences just above}}. Then the growth rate of the mutant in these patches, $a_{\text{new}}$, is not affected by the different dispersal schemes.}

Of course, there may be parameter configurations, typically high emigration rates $m$ and a bias of the wild type towards the new habitat ($\pi_w < 0$), where our assumption of density-independent reproduction is violated. Then our approximation and the numerical solution of eq.~\eqref{eq:ext_prob} in the main text strongly deviate from the simulation results (e.g. the New-New dispersal scheme in Fig.~\ref{fig:vary_m_est}(d)).
